---
title: "Review of allele and genotypic frequencies"
author: "Gota Morota"
date: "October 20, 2025"
output: 
    prettydoc::html_pretty:
      theme: architect
      highlight: vignette
      toc: true
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Overview 
In this session, we will learn how to calculate allele and genotypic frequencies in R using simulated SNP data.

## Read a file
Use the `read.table` function to import the genotype file `Geno.txt` into R as a data frame. Store the genotype data in a variable named `W`.
```{r, eval=FALSE}
W <- read.table(file = file.choose(), header=TRUE, stringsAsFactors = FALSE)
```

We can access a specific element in a data frame by specifying its coordinates within the single square bracket `[]` operator. Let’s first access the element in the first column and first row. When the row coordinate is omitted, the operator returns a data frame containing only the specified column.
```{r, eval=FALSE}
W[1, 1] # 1st row and 1st column 
head(W[, 1]) # one-column data frame
```

The following code displays the first five rows and columns of the data frame.
```{r, eval=FALSE}
W[1:5, 1:5]
```

We then drop the first column of the data frame, which contains the individual IDs. The `-` sign indicates variable removal, so `-1` means dropping the first column.
```{r, eval=FALSE}
W <- W[, -1]
```

Next, we will convert W from a data frame into a matrix. In R, matrices are generally more memory-efficient and convenient than other data types for performing linear algebra operations.
```{r, eval=FALSE}
class(W)
W <- as.matrix(W)
class(W)
```

What is the dimension of `W`?
```{r, eval=FALSE}
dim(W)
```

## Allele frequency 
Let's compute the allele frequency of the first SNP. The `table()` function returns the frequencies of genotypes.
```{r, eval=FALSE}
table(W[, 1])
```

In the first SNP, we can see that there are 100 **AA** individuals, 475 **Aa** individuals, and 429 **aa** individuals. Let's assign these numbers to variables.
```{r, eval=FALSE}
nAA <- table(W[, 1])[3]
nAa <- table(W[, 1])[2]
naa <- table(W[, 1])[1]
```

The allele frequency of *A* is given by:
$$ 
f(A) = p = \frac{2 \times (\text{no. of } AA \text{ individuals}) + 1 \times (\text{no. of } Aa \text{ individuals})}{2 \times \text{total no. of individuals}}.
$$

### Exercise 1
Use the variables `nAA`, `nAa`, and `naa` defined above to compute the allele frequencies of *A* and *a* in the first SNP.
```{r, eval=FALSE, echo=FALSE}
p <- (2 * nAA + 1 * nAa) / (2 * (nAA + nAa + naa))
p 
q <- 1 - p
q
```

## Genotypic frequency
Genotypic frequencies are given by 
$$ 
f(AA) = P = \frac{\text{No. of } AA \text{ individuals}}{\text{Total no. individuals}} \\
f(Aa) = H = \frac{\text{No. of } Aa \text{ individuals}}{\text{Total no. individuals}} \\
f(aa) = Q = \frac{\text{No. of } aa \text{ individuals}}{\text{Total no. individuals}}. \\
$$

### Exercise 2
What are the genotypic frequencies of `AA`, `Aa`, and `aa` in the first SNP? 
```{r, eval=FALSE, echo=FALSE}
P <- (nAA) / (nAA + nAa + naa)
P
H <- (nAa) / (nAA + nAa + naa)
H
Q <-  (naa) / (nAA + nAa + naa)
Q 
```

## Another approach for obtaining allele frequency is
$$ 
f(A) = p = \frac{2 \times (\text{frequency of } AA) + 1 \times (\text{frequency of } Aa)}{2 \times (\text{frequency of } AA + Aa + aa)}.
$$

### Exercise 3
Use the variables `P`, `H`, and `Q` defined above and compute the allele frequencies of *A* and *a* in the first SNP. 
```{r, eval=FALSE, echo=FALSE}
p.ex3 <- (2 * P + 1 * H) / (2 * (P + H + Q))
p.ex3 
q.ex3 <- 1 - p.ex3
q.ex3
```

### Exercise 4
What are the allele and genotypic frequencies of `AA`, `Aa`, and `aa` in the second SNP? 
```{r, eval=FALSE, echo=FALSE}
nAA <- table(W[, 2])[3]
nAa <- table(W[, 2])[2]
naa <- table(W[, 2])[1]

p <- (2 * nAA + 1 * nAa) / (2 * (nAA + nAa + naa))
p 
q <- 1 - p
q

P <- (nAA) / (nAA + nAa + naa)
P
H <- (nAa) / (nAA + nAa + naa)
H
Q <-  (naa) / (nAA + nAa + naa)
Q 
```


## Compute allele frequencies for all SNPs 
So far we have learned how to compute the allele frequency of a single SNP using the `table` function. Here, we consider how to compute allele frequencies for the entire SNPs. Of course we can apply the `table` function manually one at a time. However, this approach takes too much time to compute allele frequencies for 6,960 SNPs. Recall that allele frequency of *A* is given by 
$$ 
f(A) = p = \frac{2 \times (\text{no. of } AA \text{ individuals}) + 1 \times (\text{no. of } Aa \text{ individuals})}{2 \times \text{total no. of individuals}}.
$$
We can rewrite this equation into 
$$ 
f(A) = p = \frac{(\text{no. of } A \text{ allele in the population})}{2 \times \text{total no. of individuals}}.
$$
This suggests that all we need is the number of $A$ allele for each SNP. The `sum` function returns the number of reference allele $A$. 
```{r, eval=FALSE}
sum(W[,1]) # sum of A allele in the first SNP
sum(W[,2]) # sum of A allele in the second SNP
```

How to repeat this operation for 6,960 SNPs? The `colSums` function returns the sum of each column in a matrix as a vector. 
```{r, eval=FALSE}
colSums(W) 
```

Note that `colSums(W) ` gives the numerator of the above equation. We then divide this number by $2 \times \text{total no. of individuals}$. The function `nrows` returns the number of rows. 
```{r, eval=FALSE}
p <- colSums(W) / (2 * nrow(W))
```
The variable `p` is a vector and it contains the allele frequencies of reference allele for 6,960 SNPs. 

### Exercise 5
What is the allele frequency of reference allele in the 300th SNP?
```{r, eval=FALSE, echo=FALSE}
p[300]
```

### Exercise 6
What is the mean of reference allele frquencies in this population?
```{r, eval=FALSE, echo=FALSE}
mean(p)
```

## Minor allele frequency 
In most cases, people report a minor allele frequency, which is the frequency of less frequent allele in a given SNP. We can convert allele frequencies into minor allele frquencies by using the `ifelse` function. 
```{r, eval=FALSE}
maf <- ifelse( p > 0.5, 1-p, p)
```

### Exercise 7
What is the minor allele frquency of the reference allele in the 300th SNP?
```{r, eval=FALSE, echo=FALSE}
maf[300]
```

### Exercise 8
What is the mean of minor allele frquencies? 
```{r, eval=FALSE, echo=FALSE}
mean(maf)
```


## Visualization 
Now let's visualize the minor allele frequencies for the first 500 SNPs. 
```{r, eval=FALSE}
plot(maf[1:500])
```


















